Growth model analysis 3spp WB

### 2019-03-13 MS_Analysis_3spp_4rivers_Patterns_from 2spp.Rmd
add gams

load("D:/projects/linkedModels/data/cd_west_1997.RData")
```

```{r mean lengths by gropuingVariable}

#####################################################
#####################################################
# SElect THIS for metaPop analysis, or
#groupingVariables <- c('species','metaPop01','ageInSamples', 'year','season','age')
#groupingVariablesNoYear <- c('species','metaPop01','ageInSamples','season','age')

# THIS for 4 river analysis
groupingVariables <- c('species','riverOrdered','ageInSamples','year','season','age')
groupingVariablesNoYear <- c('species','riverOrdered','ageInSamples','season','age')

@@

riverGroupVar <- groupingVariables[2]

# Has mean size changed over time?
means <- 
  cd %>% mutate( age = year - cohort,
                 metaPop01 = ifelse(riverOrdered == 'wb obear', 0 ,1)) %>%
    filter( ageInSamples %in% 1:10, age < 3, year < 2016, area %in% c('inside','trib'), enc == 1
            #!(riverOrdered == 'wb mitchell' & ageInSamples > 8) #low sample sizes
            ) %>% 
    group_by_at(vars(one_of(groupingVariables))) %>%
    summarize(meanLen = mean(observedLength,na.rm = TRUE),
              sdLen = sd(observedLength,na.rm = TRUE),
              meanGR = mean(grLength,na.rm = TRUE),
              sdGR = sd(grLength,na.rm = TRUE),
              n = n()
              )

means$seasonGG <- factor(means$season, levels = 1:4, labels = c("Spring","Summer","Autumn","Winter"), ordered = T)
means$speciesGG <- factor(means$species, levels = c('bkt','bnt','ats'), labels = c("Brook Trout", "Brown Trout", "Atlantic Salmon"), ordered = T)

>>>>

load("D:/projects/linkedModels/data/cd_west_1997.RData")
cd <- cd %>% mutate(observedWeight = ifelse(observedWeight == -9999.9, NA, observedWeight),
                    lagObservedWeight = ifelse(lagObservedWeight == -9999.9, NA, lagObservedWeight))
```

```{r von Bert, well, gam}

vB <- 
  cd %>% mutate( age = year - cohort,
                 metaPop01 = ifelse(riverOrdered == 'wb obear', 0 ,1),
                 ydayCumul = yday + 365 * (year - cohort),
                 cf = observedWeight/observedLength^3 * 10000 ) %>%
    filter( ageInSamples %in% 1:9,
            age < 3, 
            year < 2016, area %in% c('inside','trib'), enc == 1
            #!(riverOrdered == 'wb mitchell' & ageInSamples > 8) #low sample sizes
            )

# number of knots from gam 
kValue <- 9 # from mod6 below, just for graphing

# all fish
ggplot(vB %>% filter(cohort %in% 2000:2013), aes(ydayCumul,observedLength)) +
  geom_point(size = 0.5, alpha = 0.1) +
 # geom_smooth(span = 0.1, size = 1.5, se = F) +
  geom_smooth(method='gam', alpha = 0.1, formula=y~s(x, k = kValue), se = F) +
  theme_publication()

# by species
ggplot(vB %>% filter(cohort %in% 2000:2013), aes(ydayCumul,observedLength, color = species)) +
  geom_point(size = 0.5, alpha = 0.1) +
 # geom_smooth(span = 0.1, size = 1.5, se = F) +
  geom_smooth(method='gam', alpha = 0.1, formula=y~s(x, k = kValue), se = F) +
 # facet_grid(~species) +
  theme_publication()

# by species, by cohort gridded
# ggplot(vB %>% filter(cohort %in% 2000:2013), aes(ydayCumul,observedLength, group = species, color = (species))) +
#   geom_point(size = 0.5, alpha = 0.1) +
#  # geom_smooth(span = 0.1, size = 1.5, se = F) +
#   geom_smooth(method='gam', alpha=0.1, formula=y~s(x, k = kValue), se = F) +
#   facet_wrap( ~ cohort) +
#   theme_publication()

# by species (bkt, bnt), by cohort, by river gridded
ggplot(vB %>% filter(cohort %in% 2000:2013, species %in% c('bkt', 'bnt', 'ats')), aes(ydayCumul,cf, group = species, color = (species))) +
  geom_point(size = 0.5, alpha = 0.1) +
 # geom_smooth(span = 0.1, size = 1.5, se = F) +
  geom_smooth(method='gam', alpha=0.1, formula=y~s(x, k = kValue), se = F) +
  facet_grid(riverOrdered ~ cohort) +
  theme_publication()

# gam models
library(mgcv)
mod_bnt <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr'), data = vB %>% filter(cohort %in% 2000:2013,species == 'bnt'))
summary(mod1_bnt)

mod_bkt <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr'), data = vB %>% filter(cohort %in% 2000:2013,species == 'bkt'))
summary(mod_bkt)

mod_ats <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr'), data = vB %>% filter(species == 'ats'))
summary(mod_ats)

#######################

# 1. Do we need a RE on year in s()? YES
mod0a <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr'), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod1)

mod0 <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod1)

AIC(mod0,mod0a) %>% rownames_to_column() %>% arrange(AIC)

# 2. Best structure for year,species,river
mod1 <- gam(observedLength ~ species + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod2 <- gam(observedLength ~ river + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

# river expains the most for 1-variable models
AIC(mod0,mod1,mod2) %>% rownames_to_column() %>% arrange(AIC)

# 2-way models
mod3 <- gam(observedLength ~ year * species + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod4 <- gam(observedLength ~ year * riverOrdered * riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod4)

mod5 <- gam(observedLength ~ species * riverOrdered * riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod4)

# mod5 is best
AIC(mod0,mod1,mod2,mod3,mod4,mod5) %>% rownames_to_column() %>% arrange(AIC)

# 3-way models
mod6 <- gam(observedLength ~ year * species * riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

# https://www.fromthebottomoftheheap.net/2017/12/14/difference-splines-ii/
# calculates differences from overall spline for each year
# a lot slower
vB$yearOrdered <- factor(vB$year, levels = 2000:2013)
mod6a <- gam(observedLength ~ year * species * riverOrdered + s(ydayCumul, bs = 'cr', by = (yearOrdered)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

# try with river instead of riverOrdered
mod6b <- gam(observedLength ~ year * species * river + s(ydayCumul, bs = 'cr', by = (yearOrdered)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod6b)

# 3-way is best
AIC(mod1,mod2,mod3,mod4,mod5,mod6,mod6a) %>% rownames_to_column() %>% arrange(AIC)

# Any additive 3-way better? NO
mod7 <- gam(observedLength ~ year + species * riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')))

mod8 <- gam(observedLength ~ year * species + riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')))

mod9 <- gam(observedLength ~ year + species + riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')))

AIC(mod1,mod2,mod3,mod4,mod5,mod6,mod7,mod8,mod9) %>% rownames_to_column() %>% arrange(AIC)

summary(mod6)
plot(mod6)

hist(resid(mod6))
gam.check(mod6)

# try spline on year - no real advantage, except we can see the squiggle
mod6a <- gam(cf ~ species * riverOrdered + 
                              s(ydayCumul, bs = 'cr') + 
                              s(year), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), 
             method = "REML")
plot(mod6a, scale = 0)
summary(mod6a)

```{r mean lengths by grouping Variable}

@@

# Select THIS for whole pop analysis, or
groupingVariables <- c('species','ageInSamples','year','season','age')
groupingVariablesNoYear <- c('species','ageInSamples','season','age')
riverGroupVar <- NULL

# Select THIS for metaPop analysis, or
#groupingVariables <- c('species','metaPop01','ageInSamples', 'year','season','age')
#groupingVariablesNoYear <- c('species','metaPop01','ageInSamples','season','age')
#riverGroupVar <- groupingVariables[2]

# THIS for 4 river analysis
#groupingVariables <- c('species','riverOrdered','ageInSamples','year','season','age')
#groupingVariablesNoYear <- c('species','riverOrdered','ageInSamples','season','age')
#riverGroupVar <- groupingVariables[2]
#####################################################
#####################################################

# Has mean size changed over time?
means <- 
  cd %>% mutate( age = year - cohort,
                 metaPop01 = ifelse(riverOrdered == 'wb obear', 0 ,1)) %>%
    filter( ageInSamples %in% 1:9, age < 3, year < 2016, area %in% c('inside','trib'), enc == 1
            #!(riverOrdered == 'wb mitchell' & ageInSamples > 8) #low sample sizes
            ) %>% 
    group_by_at(vars(one_of(groupingVariables))) %>%
    summarize(meanLen = mean(observedLength,na.rm = TRUE),
              sdLen = sd(observedLength,na.rm = TRUE),
              meanGR = mean(grLength,na.rm = TRUE),
              sdGR = sd(grLength,na.rm = TRUE),
              n = n()
              )

means$seasonGG <- factor(means$season, levels = 1:4, labels = c("Spring","Summer","Autumn","Winter"), ordered = T)
means$speciesGG <- factor(means$species, levels = c('bkt','bnt','ats'), labels = c("Brook Trout", "Brown Trout", "Atlantic Salmon"), ordered = T)
means$metaPop01GG <- factor(means$metaPop01, levels = c(0,1), labels = c("Brook Trout", "Brown Trout", "Atlantic Salmon"), ordered = T)
